---
title: "Consumer Complaint Database"
subtitle: "An overview of consumer complaint behavior and company responses"
author: "Luis Alarcon"
date: "October 5, 2018"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

Consumer Complaint Database
========================================================
An overview of consumer complaint behavior and company responses
-----------------------------------------------------------------
By Luis Alarcon
----------------

# Disclaimer

<p>I endeavour in this project out of my own volition. It is the first time 
that I come in contact with this data. I carry no bias before carrying this 
Exploratory Data Analysis, and the primary goal is to further my knowledge in 
the methods as well as an understanding of different areas through data 
analysis.<p>

# Introduction
## Dataset General Information

<p>The full name of the data set is "Inpatient Prospective Payment System 
(IPPS) Provider Summary for the Top 100 Diagnosis-Related Groups (DRG) - 
FY2011". It is own and updated by CMS.</p>

<p>The last update was on May 29, 2014, and the metadata was updated on 
August 30, 2017. For access to the data set as well as further information, 
please go to this link.</p>

## Dataset Contents

<p>The Dataset has 12 columns. These are the names and the description of 
them:</p>
<h7><b>This information it is copied straight from the pdf attachment found 
in this <a href = "https://data.cms.gov/api/views/97k6-zzx3/files/4496fc4f-
5f10-43e4-8183-b6da867f8981?download=true&filename=Medicare_Hospital_Inpatient_
PUF_Methodology_2017-08-30.pdf">link</a>.</b></h7>
<ol>

<li><b>DRG Definition</b></li>
    The code and description identifying the MS-DRG. MS-DRGs are a 
    classification system that groups similar clinical conditions (diagnoses) 
    and the procedures furnished by the hospital during the stay.
<li><b>Provider Id</b></li>
    The CMS Certification Number (CCN) assigned to the Medicare certified 
    hospital facility.
<li><b>Provider Name</b></li>
    The name of the provider.
<li><b>Provider Street Address</b></li>
    The provider’s street address.
<li><b>Provider City</b></li>
    The city where the provider is located.
<li><b>Provider State</b></li>
    The state where the provider is located.
<li><b>Provider Zip Code</b></li>
    The provider’s zip code.
<li><b>Provider HRR</b></li>
    The Hospital Referral Region (HRR) where the provider is located.
<li><b>Total Discharges</b></li>
    The number of discharges billed by the provider for inpatient 
    hospital services.
<li><b>Average Covered Charges</b></li>
    The provider's average charge for services covered by Medicare 
    for all discharges in the MS-DRG. These will vary from hospital to 
    hospital because of differences in hospital charge structures.
<li><b>Average Total Payments</b></li>
    The average total payments to all providers for the MS-DRG including 
    the MS-DRG amount, teaching, disproportionate share, capital, and outlier
    payments for all cases. Also included in average total payments are 
    co-payment and deductible amounts that the patient is responsible for and 
    any additional payments by third parties for coordination of benefits.
<li><b>Average Medicare Payments</b></li>
    The average amount that Medicare pays to the provider for Medicare's share 
    of the MS-DRG. Average Medicare payment amounts include the MS-DRG amount, 
    teaching, disproportionate share, capital, and outlier payments for all 
    cases. Medicare payments DO NOT include beneficiary co-payments and 
    deductible amounts norany additional payments from third parties for 
    coordination of benefits. Note: In general, Medicare FFS claims with 
    dates-of-service or dates-of-discharge on or after April 1, 2013, incurred 
    a 2 percent reduction in Medicare payment. This is in response to mandatory 
    across-the-board reductions in Federal spending, also known as 
    sequestration. For additional information, <a href= "http://www.cms.gov/Outreach-and-Education/Outreach/FFSProvPartProg/Downloads/
    2013-03-08-standalone.pdf">visit</a>. 
</ol>

<p>The initial process of refining the dataset was done in a python script 
named "u_project_EDA-CMS_python3.ipynb", the code is presented in 
"u_project_EDA-CMS_python3.html"; for further details examine the HTML file.</p>

# Exploratory Data Analysis (EDA)
## EDA: Initial Data Exploration
### Data Summary

<p>Firstly, we load the packages that we are going to use in this project.</p>
```{r echo=FALSE, warning=FALSE, message=FALSE} 
#The following code maanges the output of messages and warning when the code is run. The code also manages the figure size of an image created in the project.
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)
#The libraries used in this project are loaded
library(ggplot2)
library(gridExtra)
library(grid)
library(lattice)
library(GGally)
library(dplyr) 
```

<p>Secondly, we load the dataset from its original location to the 
data frame df_cms.</p>

```{r echo=FALSE}
#We load the dataset and produce a first look of the data containing it.
df_cms <- read.csv("./datasets/df_cms.csv")
head(df_cms)
```


<p>Before we continue, we check if there are any null values within the 
data frame df_cms.</p>

```{r echo=FALSE}
#We check if there are any nulls in the dataset
df_null <- df_cms[is.na(df_cms),]

print(head(df_null))
```

<p>There are no null values in the data frame. Afterwards, we check the 
number of entries present in this dataset.</p>

```{r echo=FALSE}
#We check the dimensions of the dataset
dim(df_cms)
```

<p>The number of entries is 163065. The number of entries is significant. 
Therefore, We proceed to make a quick summary of the data.</p>

```{r echo=FALSE}
#We produce a summary of the dataset
summary(df_cms)
```


<p>Due to the vast array of data and groups found in this particular 
data frame. We will begin our analysis per regions. We will start with 
"Census Region" since it encompasses all states within the four central 
regions of the USA, the divisions within regions, the Federal bank that is in 
charge of the state as well as the Regions established by the Bureau of 
Economic Analysis.</p>

### Univariate Plots Section

<p>This graph contains the number of entry per each of the groups mentioned 
above.</p>

```{r echo=FALSE, Bivariate_Plots,fig.width=9,fig.height=10}
# We will produce our first graph of the dataset
#P1 produces a bar graph with the count of entries 
#that are from the same census region
p1 = ggplot(df_cms) + 
  geom_histogram(aes(x = Census.Region, fill = Census.Region), stat = "count") + 
  ggtitle("Number of CMS per Census Region.")+
  scale_x_discrete(breaks = 1:4, labels=c("MIDWEST","NORTHEAST","SOUTH","WEST")) +
  xlab(NULL)
#P2 produces a bar graph with the count of entries 
#that are from the same census region division
p2 = ggplot(df_cms) + geom_histogram(aes(x = Census.Region.Division, 
                                         fill = Census.Region.Division), 
                                     stat = "count") + 
  ggtitle("Number of CMS per Census Region Division.") +
  scale_x_discrete(breaks = 1:9, labels=c("D1","D2","D3","D4","D5","D6",
                                          "D7","D8","D9")) +
  xlab(NULL)
#P3 produces a bar graph with the count of entries 
#that are from the same federal region
p3 = ggplot(df_cms) + geom_histogram(aes(x = Federal.Region, 
                                         fill = Federal.Region),
                                     stat = "count") + 
  ggtitle("Number of CMS per Federal Bank Region.")+
  scale_x_discrete(breaks = 1:9, labels=c("REGION_I","REGION_II","REGION_III",
                                          "REGION_IV","REGION_V","REGION_VI",
                                          "REGION_VII",
                                          "REGION_VIII","REGION_X")) +
  xlab(NULL)
#P4 produces a bar graph with the count of entries 
#that are from the same economic analysis region
p4 = ggplot(df_cms) + geom_histogram(aes(x = Economic.Analysis.Region, 
                                         fill = Economic.Analysis.Region), 
                                     stat = "count") + 
  ggtitle("Number of CMS per Bureau of Economic Analysis Region.")+
  scale_x_discrete(breaks = 1:8, labels=c("D1","D2","D3","D4","D5","D6",
                                          "D7","D8")) +
  xlab(NULL)
#P5 produces a bar graph with the count of entries 
#that are from the same census region with legends of the census region division
p5 = ggplot(aes(x=Census.Region, fill = Census.Region.Division), data= df_cms, 
            stat = "count")+
  geom_bar()+ 
  ggtitle("Number of CMS per Census Region with the divisions shown.")+
  scale_x_discrete(breaks = 1:4, labels=c("MIDWEST","NORTHEAST",
                                          "SOUTH","WEST")) +
  xlab(NULL)

lay <- rbind(c(1,1),
             c(2,3),
             c(4,5))
grid.arrange(p1, p2, p3, p4, p5, layout_matrix = lay)
```

### Univariate Analysis Section

<p>We will start the initial analysis by obtaining the frequency for each 
one of the variables of the columns:</p> 

<ol>
<li>"Census.Region"</p>
<li>"Census Region Division"</p>
<li>"Federal Region"</p>
<li>"Bureau of Economic Analysis Region"</p>
</ol>

```{r}
#we load the package plyr
library(plyr)
#We create tables shown the frequencies shown in the images shown above.
print(count(df_cms, vars = "Census.Region"))

print(count(df_cms, vars = "Census.Region.Division"))

print(count(df_cms, vars = "Federal.Region"))

print(count(df_cms, vars = "Economic.Analysis.Region"))

```

<p>In the Census Region, the region South has the highest account of 
entries with 67038 and the lowest in the West with 26710. In the Divisions 
in the Census Region, the Division 5 South Atlantic has the hights entries 
with 34116, and the lowest in the Division 1 New England.</p>

<p>When considering Regions under the different Federal Banks, Region IV is 
the one with the highest entry and the region with the lowest entries is 
Region_XIII with 4283. Lastly, The Rocky Mountain Region of the Bureau of 
Economic Analysis is the region with the lowest entries in 3926. On the other 
hand, Southeast is the region with the highest entries with 48468.</p>

<p>Before we continue, we will get the summary of the Total Discharges, 
Average Covered Charges, Average Total Payments and Average Medicare 
Payments per Census Region.</p>

```{r}
#We generate a summary description  of the Total Discharges by regions
by(df_cms$Total.Discharges, df_cms$Census.Region, summary)
```


```{r}
#We generate a summary description  of the Average Covered Charges by regions
by(df_cms$Average.Covered.Charges, df_cms$Census.Region, summary)
```

```{r}
#We generate a summary description  of the Average Total Payments by regions
by(df_cms$Average.Total.Payments, df_cms$Census.Region, summary)
```


```{r}
#We generate a summary description  of the Average Medicare Payments by regions
by(df_cms$Average.Medicare.Payments, df_cms$Census.Region, summary)
```

## EDA: Furthe Data Exploration
### Bivariate Plots Section
#### Bivariate Plots Section: Plots
<p>In order to continue this analysis, we will proceed to see the relationship 
between Total Discharges, Averaged Covered Charges (USD), Average Total Payments
(USD) and Average Medicare Payments (USD).</p>

```{r echo=FALSE, Multivariate_Data_Plots,fig.width=9,fig.height=10}
theme_set(theme_minimal(12))

# set the seed for reproducible results
set.seed(2000)
df_cms_subset <- df_cms[,c("Total.Discharges","Average.Covered.Charges",
                           "Average.Total.Payments",
                           "Average.Medicare.Payments")]
ggpairs(df_cms_subset[sample.int(nrow(df_cms_subset),2000), ])
```

<p>From the graph, we can see that the relationship between Total Discharges 
and Average Covered Charges is that the fewer discharges the more Average 
Covered Charges. This trend is the same when we compare Total Discharges is 
plotted alongside Average Total Payments and Average Medicare Payments.</p>

<p>When a correlation test is performed between Total Discharges and the other 
variables these are the results:</p>
<ul>
<li><b>Total Discharges vs Average Covered Charges(USD)</b>:</li> Corr: -0.025
<li><b>Total Discharges vs Average Total Payments(USD)</b>:</li> Corr: 0.0151
<li><b>Total DIscharges vs Average Medicare Payments(USD)</b>:</li> Corr:0.0121
</ul>

<p>The results indicate that there is no correlation between the number of 
discharges and the payments made or received by patients or Medicare.</p>

<p>The graph also shows the relationship between Average Covered Charges(USD) 
and Average Total Payments(USD) and Average Medicare Payments(USD). The graph 
shows almost a more linear positive relationship between Average Covered 
Charges(USD) and Average Total Payments(USD) and Average Medicare Payments 
(USD). </p>

<p>The graph is supported by the data extracted in the correlation test between 
the variables:</p>

<li><b>Average Covered Charges(USD) vs Average Total Payments(USD)</b>:</li> 
Corr: 0.772
<li><b>Average Covered Charges(USD) vs Average Medicare Payments(USD)</b>:</li> 
Corr: 0.764

<p>The results indicate that there is a positive relationship between the 
variable and due to the similarity in results it further supports the 
similarity between the two relationships.</p>

</p>The last relationship that could be explored from the graph is between 
Average Total Payments(USD) and Average  Medicare Payments(USD). The graph 
shows a very linear relationship between the two variables, and this is further 
supported by the correlation test showing a Corr value of 0.992.</p>


### Multivariable Section
#### Multivariable Section: Plots

<p>We will start by analysing the following group of variables together:</p>
<ol>
<li>Average Covered Charges vs Average Medicare Payments: This comparison was 
chosen because Average Covered Charges are the average charge for services 
covered by Medicare for all discharges in the-DRG and Average Medicare Payments 
are the average amount that Medicare pays providers for Medicare's share of the 
MS-DRG.</li>

<li>Total Discharges vs Average Total Payments: This comparison was chosen 
because the total discharges account for the number of discharges billed by 
the provider for inpatient hospital services and the average total payments 
account for the total the average total payments to all providers for the 
MS-DRG; including partial coverage by the patient.</li>

<li>Average Total Payments vs Average Medicare Payments.</li>
</ol>

##### Average Covered Charges vs Average Medicare Payments

###### CC-vs-MP:Plots

<p>The first graph that is going to show the whole data from df_cms. 
The whole data is graphed in this way in order to acquire the whole pattern of 
the data between the two variables. </p>

```{r echo=FALSE, Bivariate_Plots_Total_CC-vs-MP,fig.width=9,fig.height=11}
#We create a scatter plot between 
#the average covered charges vs average medicare payments
ggplot(aes(x=Average.Covered.Charges, y=Average.Medicare.Payments, 
           color=Census.Region),data= df_cms) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0))+ 
  ggtitle("Average Covered Charges vs Average Medicare Payments")
```

<p>As seen on the graph, the majority of the data is under the x-value 
250000&#36; and y-value 50000&#36;. Due to this the second graph will have the limits 
mentioned above and also 4 further graphs will be added; the entire data 
frame data selectively gather under the four census regions of the USA, 
Midwest, Northeast, South and West.</p>

```{r echo=FALSE, Bivariate_Plots_CR_CC-vs-MP,fig.width=9,fig.height=11}
#We create a series of plots that showcase 
#the average covered charges vs average medicare payments
#This graph show an overall view of all regions
p1 =ggplot(aes(x=Average.Covered.Charges, y=Average.Medicare.Payments, 
               color=Census.Region),data= df_cms) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0))+ 
  ggtitle("Average Covered Charges vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,250000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#average covered charges vs average medicare payments from the Midwest
p2 = ggplot(aes(x=Average.Covered.Charges, y=Average.Medicare.Payments),
                data=subset(df_cms, Census.Region == "MIDWEST")) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0), color = "blue")+ 
  ggtitle("Midwest Region: Average Covered Charges 
          vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,250000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#average covered charges vs average medicare payments from the Northeast
p3 = ggplot(aes(x=Average.Covered.Charges, y=Average.Medicare.Payments),
                data=subset(df_cms, Census.Region == "NORTHEAST")) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0), color = "green")+ 
  ggtitle("Northeast: Average Covered Charges vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,250000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#average covered charges vs average medicare payments from the South
p4 = ggplot(aes(x=Average.Covered.Charges, y=Average.Medicare.Payments),
                data=subset(df_cms, Census.Region == "SOUTH")) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0), color = "red")+ 
  ggtitle("South: Average Covered Charges vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,250000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#average covered charges vs average medicare payments from the West
p5 = ggplot(aes(x=Average.Covered.Charges, y=Average.Medicare.Payments),
                data=subset(df_cms, Census.Region == "WEST")) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0), color = "brown")+ 
  ggtitle("West: Average Covered Charges vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,250000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 

#plotting the graphs
lay <- rbind(c(1,1),
             c(2,3),
             c(4,5))
grid.arrange(p1, p2, p3, p4, p5, layout_matrix = lay)
```

<p>As we can see from the graph West and Midwest regions, have a similar 
scatter plot, flatter and more left-leaning. However, Midwest has a more 
linear shape than the West region.</p>

<p>Northeast and South regions have similar scatterplots with widening x 
and y profile than other regions. However, the South region have a straight 
line. It will be interesting to know why it was developed.</p>

###### CC-vs-MP:Analysis

<p>We start the analysis by determining the Correlation between the 
two variables. We will perform a Pearson test between the two variables 
without subdividing the data into regions.</p>
 
```{r}
#We do a correlation test between 
#Average Covered Charges and Average Medicare Payments
res <- cor.test(df_cms$Average.Covered.Charges, 
                df_cms$Average.Medicare.Payments, 
                    method = "pearson")

res
```

<p>The test shows a cor value of 0.767. Values between 0.7 and 0.9 state that 
there is a steady uphill linear relationship between Average Covered Charges 
and Average Medicare payments.</p>

<p>For the moment we will only explore the correlation between the two 
variables, and in a subsequent analysis with more detail and in-depth 
data we will explore other analysis between these two variables</p>

##### Total Discharges vs Average Total Payments

###### TD-vs-TP:Plots

<p>This first graph is going to deliver an overall picture of the relation 
between the two data columns; including data patterns.</p>

```{r echo=FALSE, Bivariate_Plots_Total_TD-vs-TP,fig.width=9,fig.height=11}
#We create a scatter plot between 
#the total discharges vs average total payments
ggplot(aes(x=Total.Discharges, y=Average.Total.Payments, 
           color=Census.Region),data= df_cms) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0))+ 
  ggtitle("Average Total Discharges vs Average Total Payments")
```

<p>This graph shows that the majority of total discharges are fewer than 
250 with an increase average total payments are mostly under 25000&#36;.</p>

<p>The next graph will plot the same graph as above but with limits of 
x-value 500 and y-value 50000&#36;, as well as four separate graphs with the 
data frame gather selectively from the regions of the USA.</p>

```{r echo=FALSE, Bivariate_Plots_CR_TD-vs-TP,fig.width=9,fig.height=11}
#We create a series of plots that showcase the total discharges vs 
#average total payments
#This graph show an overall view of all regions
p1 =ggplot(aes(x=Total.Discharges, y=Average.Total.Payments, 
               color=Census.Region),data= df_cms) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0))+ 
  ggtitle("Total Discharges vs Average Total Payments")+
  ylim(0,50000)+
  xlim(0,500)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#total discharges vs average total payments from the Midwest
p2 = ggplot(aes(x=Total.Discharges, y=Average.Total.Payments),
            data=subset(df_cms, Census.Region == "MIDWEST")) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0), color = "blue")+ 
  ggtitle("Midwest Region: Total Discharges vs Average Total Payments")+
  ylim(0,50000)+
  xlim(0,500)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#total discharges vs average total payments from the Northeast
p3 = ggplot(aes(x=Total.Discharges, y=Average.Total.Payments),
            data=subset(df_cms, Census.Region == "NORTHEAST")) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0), color = "green")+ 
  ggtitle("Northeast: Total Discharges vs Average Total Payments")+
  ylim(0,50000)+
  xlim(0,500)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#total discharges vs average total payments from the South
p4 = ggplot(aes(x=Total.Discharges, y=Average.Total.Payments),
            data=subset(df_cms, Census.Region == "SOUTH")) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0), color = "red")+ 
  ggtitle("South: Total Discharges vs Average Total Payments")+
  ylim(0,50000)+
  xlim(0,500)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#total discharges vs average total payments from the West
p5 = ggplot(aes(x=Total.Discharges, y=Average.Total.Payments),
            data=subset(df_cms, Census.Region == "WEST")) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0), color = "brown")+ 
  ggtitle("West: Total Discharges vs Average Total Payments")+
  ylim(0,50000)+
  xlim(0,500)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 

#plotting the graphs
lay <- rbind(c(1,1),
             c(2,3),
             c(4,5))
grid.arrange(p1, p2, p3, p4, p5, layout_matrix = lay)
```

<p>All graphs share a similar pattern than the one shared from the pattern 
mention in the general and global graph. There are two distinct protrusions 
of data reflecting an increase of total discharges with approximately an 
average total payments of about 10000&#36;; with a constant increase of 
discharges, similar to a horizontal line. The pattern is repeated about 
15000&#36;  Average Total Payments; it is less apparent in the West Region.</p>

###### TP-vs-MP:Analysis

<p>We start the analysis by determining the Correlation between the two 
variables. We will perform a Pearson test between the two variables without 
subdividing the data into regions.</p>

```{r}
#We do a correlation test between 
#Total Discharges vs Average Total Payments
res <- cor.test(df_cms$Total.Discharges, df_cms$Average.Total.Payments, 
                    method = "pearson")

res
```

<p>The test shows a cor value of -0.016. Values between 0 and -0.3 state that 
there is a weak negative linear relationship between Total Discharges and 
Average Total Payments.</p>

<p>It is unusual that this the correlation found between these variables; 
especially since these two variables involve the total of discharges and 
average total payments.</p>  

<p>For the moment we will only explore the correlation between the two 
variables, and in a subsequent analysis with more detail and in-depth data, 
we will explore other analysis between these two variables.</p>


##### Average Total Payments vs Average Medicare Payments

###### TP-vs-MP:Plots

<p>This first graph is going to deliver an overall picture of the relation 
between the two data columns; including data patterns.</p>

```{r echo=FALSE, Bivariate_Plots_Total_TP-vs-MP,fig.width=9,fig.height=11}
#We do a general graph between 
#Average Total Payments and Average Medicare Payments
ggplot(aes(x=Average.Total.Payments, y=Average.Medicare.Payments, 
           color=Census.Region),data= df_cms) + 
  geom_point(alpha = 1/5, position = position_jitter(h=0))+ 
  ggtitle("Average Total Payments vs Average Medicare Payments")
```

<p>This graph shows a linear-like pattern. The majority of the entries are 
within the liner pattern. Also, the majority of entries are within x-value 
and y-value limits of 50000&#36;.</p>

```{r echo=FALSE, Bivariate_Plots_CR_TP-vs-MP,fig.width=9,fig.height=11}
#We create a series of plots that showcase 
#the Average Total Payments and Average Medicare Payments
#This graph show an overall view of all regions
p1 =ggplot(aes(x=Average.Total.Payments, y=Average.Medicare.Payments, 
               color=Census.Region),data= df_cms) + 
  geom_point(alpha = 1/40, position = position_jitter(h=0))+ 
  ggtitle("Average Total Payments vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,50000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#Average Total Payments and Average Medicare Payments from the Midwest
p2 = ggplot(aes(x=Average.Total.Payments, y=Average.Medicare.Payments),
            data=subset(df_cms, Census.Region == "MIDWEST")) + 
  geom_point(alpha = 1/40, position = position_jitter(h=0), color = "blue")+ 
  ggtitle("Midwest Region: Average Total Payments vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,50000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#Average Total Payments and Average Medicare Payments from the Northeast
p3 = ggplot(aes(x=Average.Total.Payments, y=Average.Medicare.Payments),
            data=subset(df_cms, Census.Region == "NORTHEAST")) + 
  geom_point(alpha = 1/40, position = position_jitter(h=0), color = "green")+ 
  ggtitle("Northeast: Average Total Payments vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,50000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#Average Total Payments and Average Medicare Payments from the South
p4 = ggplot(aes(x=Average.Total.Payments, y=Average.Medicare.Payments),
            data=subset(df_cms, Census.Region == "SOUTH")) + 
  geom_point(alpha = 1/40, position = position_jitter(h=0), color = "red")+ 
  ggtitle("South: Average Total Payments vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,50000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 
#This graph shows a scatter plot between 
#Average Total Payments and Average Medicare Payments from the West
p5 = ggplot(aes(x=Average.Total.Payments, y=Average.Medicare.Payments),
            data=subset(df_cms, Census.Region == "WEST")) + 
  geom_point(alpha = 1/40, position = position_jitter(h=0), color = "brown")+ 
  ggtitle("West: Average Total Payments vs Average Medicare Payments")+
  ylim(0,50000)+
  xlim(0,50000)+
  theme(text = element_text(size=9),plot.title = element_text(size=8), 
        axis.text=element_text(size=6),plot.margin=unit(c(1,1,1,1), "cm")) 

#plotting the graphs
lay <- rbind(c(1,1),
             c(2,3),
             c(4,5))
grid.arrange(p1, p2, p3, p4, p5, layout_matrix = lay)
```

<p>All graphs show the same pattern. There are mild differences in the width. 
Further analysis could reveal more about this pattern.</p>

###### TP-vs-MP:Analysis

<p>We start the analysis by determining the Correlation between the two 
variables. We will perform a Pearson test between the two variables without 
subdividing the data into regions.</p>

```{r}
#A correlation test is made between 
#Average Total Payments and Average Medicare Payments
res <- cor.test(df_cms$Average.Total.Payments, df_cms$Average.Medicare.Payments, 
                    method = "pearson")

res
```

<p>The test shows a cor value of 0.989. Values between 0.7 and 0.9 state that 
there is a steady uphill linear relationship between Total Payments and Average 
Medicare Payments.</p>

<p>For the moment we will only explore the correlation between the two 
variables, and in a subsequent analysis with more detail and in-depth data, 
we will explore other analysis between these two variables.</p>









```{r echo=FALSE, total_discharges_vs_census_region_division,fig.width=9,fig.height=9}
ggplot(aes(x=Census.Region.Division, fill=Census.Region),data= df_cms)+
  geom_boxplot(aes(y= Total.Discharges))+
  coord_cartesian(ylim = c(0, 90))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

<p>In the region of the Northeast of the USA, we see that the Interquartile 
Range is between ~18 for the 25% percentile and ~67 for the 75% percentile.</p>

<p>In the region of the Midwest of the USA, we see that the Interquartile 
Range is between ~18 for the 25% percentile and ~67 for the 75% percentile.</p>

<p>In the region of the South of the USA, we see that the Interquartile 
Range is between ~18 for the 25% percentile and ~50 for the 75% percentile.</p>

<p>In the region of the West of the USA, we see that the Interquartile 
Range is between ~15 for the 25% percentile and ~42 for the 75% percentile.</p>


### Final Plots and Summary 

<p>This section only handles aesthetic modification of these images. 
The three images chosen for this section are:</p>

<ol>
  <li>Average Covered Charges vs Average Medicare Payments</li>
  <li>Average Total Payments vs Average Medicare Payments.</li>
  <li>Total Discharges by Census Region Division.</li>
<ol>

```{r echo=FALSE, Bivariate_Plots_Total_CC-vs-MP_final,fig.width=7,fig.height=7}
#We create a scatter plot between 
#the average covered charges vs average medicare payments
ggplot(aes(x=Average.Covered.Charges, y=Average.Medicare.Payments, 
           color=Census.Region),data= df_cms) + 
  geom_point(alpha = 1/3, position = position_jitter(h=0))+ 
  ggtitle("Average Covered Charges vs Average Medicare Payments")+
  guides(color=guide_legend(title="Census\nRegion", 
                            override.aes = list(alpha = 1)))+
  xlab("Average Covered Charges ($)")+
  ylab("Average Medicare Payemnts ($)")
```



```{r echo=FALSE, Bivariate_Plots_Total_TP-vs-MP_Final,fig.width=7,fig.height=7}
#We do a general graph between 
#Average Total Payments and Average Medicare Payments
ggplot(aes(x=Average.Total.Payments, y=Average.Medicare.Payments, 
           color=Census.Region),data= df_cms) + 
  geom_point(alpha = 1/3, position = position_jitter(h=0))+ 
  ggtitle("Average Total Payments Average Medicare Payments")+
  guides(color=guide_legend(title="Census\nRegion", 
                            override.aes = list(alpha = 1)))+
  xlab("Average Total Payments ($)")+
  ylab("Average Medicare Payemnts ($)")
```


```{r echo=FALSE, total_discharges_vs_census_region_division_final,fig.width=7,fig.height=7}
ggplot(aes(x=Census.Region.Division, fill=Census.Region),data= df_cms)+
  geom_boxplot(aes(y= Total.Discharges))+
  coord_cartesian(ylim = c(0, 90))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  guides(color=guide_legend(title="Census\nRegion", 
                            override.aes = list(alpha = 1)))+
  xlab("Census Region Division")+
  ylab("Total Discharges")+ 
  ggtitle("Total Discharges by Census Region Division")+ 
  scale_x_discrete(labels=c("D1_NEW_ENGLAND" = "New England", 
                            "D2_MID-ATLANTIC" = "Mid Atlantic",
                            "D3_EAST_NORTH_CENTRAL" = "East North Central",
                            "D4_WEST_NORTH_CENTRAL" = "West North Central",
                            "D5_SOUTH_ATLANTIC" = "South Atlantic",
                            "D6_EAST_SOUTH_CENTRAL" = "East South Central",
                            "D7_WEST_SOUTH_CENTRAL" = "West South Central",
                            "D8_MOUNTAIN" = "Mountain","D9_PACIFIC" = "Pacific"))
```



# Reflection
<p>The Dataset is one that shows a significant amount of data towards Medicare 
payment information in hospitals throughout the USA.  It consists of 16 
variables and 163065 entries. We explore the various relationships between 
total payments, Medicare payments, Total Discharges, and regions within the 
USA.</p>

<p>The dataset is vast and much more analysis could be performed, for example, 
the differences in payment from Medicare and Total payments made per procedure 
per hospitals in the same region as well as through states or regions in the 
USA. </p>

</p>The vastness of this dataset it is on itself a struggle, the analysis made 
in here it is mainly superficial and many more aspects could be analysed 
further. However, this analysis does give an overview picture of the 
distribution of payments and discharges stored and how that is represented in 
the different regions of the USA.<p>

<p>Furthermore, another difficulty obtain was to allocate the data into regions, 
divisions and so on. No python package is readily available that provides the 
information required, therefore it needed to be done by hand and checked in a 
multitude of sources, so the information was accurate.</p>

<p>One question that could further be asked form this dataset is how the 
different medical procedures are between payments(All the different types of 
payment related variables in this dataset) and discharges among states, regions 
and divisions. It is understood that different hospitals have a different 
payment structure or charge structure per procedure, so a further analysis 
could show this pattern and how much more significant such a difference is.</p>



